On the analytic structure of the linear response susceptibility in open systems

### 2018-03-09 AnalyticStructure_main.tex
update

The response density operator is expressed as the first-order variation of ground state density $\dmnot = \sum_{\{p\}} \ket{\psi_p} \bra{\psi_p}$
%\be\label{rhodef1}
%\dmnot = \sum_{\{p\}} \ket{\psi_p} \bra{\psi_p}  \;,
%\ee
where the set $\{\ket{\psi_p}\}$ denote the occupied states, \emph{i.e} the subset of the bound eigenstates of $\hnot$ with energy $\eps_p$ lower than the Fermi level.
By considering the linear order time evolution of these states in presence of the perturbation we can derive the equation of motion of the response density in the form of 
a quantum Liouville equation (for example see \cite{baroni2008}). This procedure suggests us to introduce an explicit representation of the response density in terms of a set of 
$\omega$-dependent \emph{fluctuation states} $\{\ket{f_p(\omega)}\}$, as follows: 
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p(-\omega)} + \ketbra{f_p(\omega)}{\psi_p}\right) \;.
\ee
An analysis of the Liouville equation for the response density written in this fashion evidences that fluctuation states are orthogonal to the occupied eigenstates 
of $\hnot$ and their equations of motion are written in the form of a modified Sternheimer equation \cite{mahan1980}:
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'(\omega))\ket{\psi_p} \;,
\ee
where $\op Q_0=\identity-\dmnot$ is the projector in the subspace orthogonal to the occupied states and the operator $\op V'(\omega)$ is a perturbing term due to the density-dependence 
of $\hnot$. %the unperturbed Hamiltonian. 
%and describes the perturbation induced by the variation of $\hnot$ under the action of the perturbing field.
%The  r.h.s of \eqref{fluctuationStateEqMotion1} expresses the complete perturbing potential

The interesting aspect of equations \eqref{fluctuationStateEqMotion1} in the perspective of the present analysis is that they allow us to present a 
\emph{localization argument} for the fluctuation states. To clarify this statement we split the unperturbed Hamiltonian as the sum of the kinetic $\op T$ and potential 
$\op V$ term and solve \eqref{fluctuationStateEqMotion1} in the form of a self-consistent inhomogeneous Helmholtz equation by writing:
\be
\ket{f_p(\omega)} = \op G_H(\omega+\epsilon_p)\ket{s(\omega)} \;,
\ee
where we have introduced the Green function of the Helmholtz operator $\op G_H(\omega) = (\omega-\op T)^{-1}$, % with $k^2_p(\omega) = (\omega+\eps_p)$,
and the \emph{source term} is given by:
\be
\ket{s(\omega)} =  \op V\ket{f_p(\omega)} + \op Q_0(\op V'(\omega) + \op \Phi(\omega))\ket{\psi_p} \;.\nn
\ee
For open boundary conditions in three dimensions, the kernel of $\op G_H(\omega)$ in the coordinate representation reads: 
% \be\lb{HelmholtzKernelDef1}
% \bra{\r}\op G_H(\omega)\ket{\r'} = \frac{1}{4\pi} \left\{ 
% \begin{matrix}
% \frac{e^{-2k_p(\omega)|\r-\r'|}}{|\r-\r'|} \;,\; {\textrm for} \; \omega \leq |\eps_p| \\ 
% \frac{e^{\ii k_p(\omega)|\r-\r'|}}{|\r-\r'|} \;,\; {\textrm for} \; \omega > |\eps_p|
% \end{matrix}
% \right. \;,
% \ee
\be\lb{HelmholtzKernelDef1}
\bra{\r'} \op G_H(k^2) \ket\r = \frac{1}{4\pi} \begin{cases}
\frac{e^{-2k|\r - \r' |}}{|\r- \r' |} \;,\; & {\textrm for}\; k < 0\; \\ 
\frac{e^{2\ii k |\r-\r' |}}{|\r-\r' |} \;, & {\textrm for} \; k > 0
\end{cases}
\ee
%and the $\r$-representation of the fluctuation states $f_p(\omega,\r)$ is expressed through a convolution integral between the kernel \eqref{HelmholtzKernelDef1} and
%$\brket{\r}{s(\omega)}$.
An inspection of the structure of \eqref{HelmholtzKernelDef1} evidences that, for each value of $p$, there exist a threshold value $\omega = |\eps_p|$, 
which govern the localization properties of the corresponding fluctuation state. Indeed, for $\omega > |\eps_p|$,  $G_H(\omega,\r)$ behaves as a spherical wave and its
convolution with the source term always give raises to a $f_p(\omega,\r)$ delocalized in the whole accessible space. On the contrary, for values of $\omega$ below the threshold,
$G_H(\omega,\r)$ contains an exponential damping factor and the localization features of $f_p(\omega,\r)$ are determined by the spatial behavior of the source term.

@@

Numerical example (convergence below threshold of the Benzene spectrum computed in different computational setups).

>>>>

The response density operator is expressed as the first-order variation of ground state density $\dmnot = \sum_{\{p\}} \ket{\psi_p} \bra{\psi_p}$
%\be\label{rhodef1}
%\dmnot = \sum_{\{p\}} \ket{\psi_p} \bra{\psi_p}  \;,
%\ee
where the set $\{\ket{\psi_p}\}$ denote the occupied states, \emph{i.e} the subset of the bound eigenstates of $\hnot$ with energy $\eps_p$ lower than the Fermi level.
By considering the linear order time evolution of these states in presence of the perturbation we can derive the equation of motion of the response density in the form of 
a quantum Liouville equation (for example see \cite{baroni2008}). This procedure suggests us to introduce an explicit representation of the response density in terms of a set of 
$\omega$-dependent \emph{fluctuation states} $\{\ket{f_p(\omega)}\}$, as follows: 
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p(-\omega)} + \ketbra{f_p(\omega)}{\psi_p}\right) \;.
\ee
Fluctuation states are, by construction, orthogonal to the occupied eigenstates of $\hnot$. Moreover, an analysis of the Liouville equation for the response density written in this fashion
evidences that their equations of motion are written in the form of a modified Sternheimer equation \cite{mahan1980}:
%An analysis of the Liouville equation for the response density written in this fashion evidences that fluctuation states are orthogonal to the occupied eigenstates 
%of $\hnot$ and their equations of motion are written in the form of a modified Sternheimer equation \cite{mahan1980}:
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'(\omega))\ket{\psi_p} \;,
\ee
where $\op Q_0=\identity-\dmnot$ is the projector in the subspace orthogonal to the occupied states of $\hnot$ and the operator $\op V'(\omega)$ is a perturbing term due to the density-dependence 
of the unperturbed Hamiltonian. %$\hnot$. %the unperturbed Hamiltonian. 
%and describes the perturbation induced by the variation of $\hnot$ under the action of the perturbing field.
%The  r.h.s of \eqref{fluctuationStateEqMotion1} expresses the complete perturbing potential

The interesting aspect of equations \eqref{fluctuationStateEqMotion1} in the perspective of the present analysis is that they allow us to present a 
\emph{localization argument} for the fluctuation states. To clarify this statement we split the unperturbed Hamiltonian as the sum of the kinetic $\op T$ and potential 
$\op V$ term and solve \eqref{fluctuationStateEqMotion1} in the form of a self-consistent inhomogeneous Helmholtz equation by writing:
\be
\ket{f_p(\omega)} = \op G_H(\omega+\epsilon_p)\ket{s(\omega)} \;,
\ee
where we have introduced the Green function of the Helmholtz operator $\op G_H(\omega) = (\omega-\op T)^{-1}$, % with $k^2_p(\omega) = (\omega+\eps_p)$,
and the \emph{source term} is given by:
\be
\ket{s(\omega)} =  \op V\ket{f_p(\omega)} + \op Q_0(\op V'(\omega) + \op \Phi(\omega))\ket{\psi_p} \;.\nn
\ee
For open boundary conditions in three dimensions, the kernel of $\op G_H(\omega)$ in the coordinate representation reads: 
% \be\lb{HelmholtzKernelDef1}
% \bra{\r}\op G_H(\omega)\ket{\r'} = \frac{1}{4\pi} \left\{ 
% \begin{matrix}
% \frac{e^{-2k_p(\omega)|\r-\r'|}}{|\r-\r'|} \;,\; {\textrm for} \; \omega \leq |\eps_p| \\ 
% \frac{e^{\ii k_p(\omega)|\r-\r'|}}{|\r-\r'|} \;,\; {\textrm for} \; \omega > |\eps_p|
% \end{matrix}
% \right. \;,
% \ee
\be\lb{HelmholtzKernelDef1}
\bra{\r} \op G_H(k^2) \ket{\r'} = \frac{1}{4\pi} \begin{cases}
\frac{e^{2k|\r - \r' |}}{|\r- \r' |} \;,\; & {\textrm for}\; k < 0\; \\ 
\frac{e^{2\ii k |\r-\r' |}}{|\r-\r' |} \;, & {\textrm for} \; k > 0
\end{cases} \;,
\ee
and the $\r$-representation of the fluctuation states $f_p(\omega,\r)$ is expressed through a convolution integral between the kernel \eqref{HelmholtzKernelDef1} and
$\brket{\r'}{s(\omega)}$.
An inspection of the structure of \eqref{HelmholtzKernelDef1} evidences that, for each value of $p$, there exist a threshold value $\omega = |\eps_p|$, 
which govern the localization properties of the corresponding fluctuation state. Indeed, for $\omega > |\eps_p|$,  $G_H(\omega)$ behaves as a spherical wave and its
convolution with the source term always give raises to a $f_p(\omega,\r)$ delocalized in the whole accessible space. On the contrary, for values of $\omega$ below the threshold,
the kernel contains an exponential damping factor and the localization features of $f_p(\omega,\r)$ are determined by the spatial behavior of the source term.

\vspace{0.3cm}
\noindent{TODO: Add comment to support the localization below threshold}. Two possible way. First way, use the fact that the $\ket{f_p(0)}$ is the nothing but the $p$-\emph{th} 
eigenstate of the perturbed Hamiltonian with perturbing field $\op\Phi(0)$. The localization of $\ket{f_p(0)}$ can be explicitly checked by solving the ground-state problem for
this Hamiltonian. How to extended to finite values of $\omega$ below threshold? Second way. Try and iterative solution of the self-consistent Helmholtz equation (below threshold) 
starting from a localized guess for the fluctuation state. Then state something about the the fact that is reasonable to believe that (for a proper choice of the perturbing field)
the operators appearing in the source term keep its support compact. If we are able to state this in a appropriate way the localization argument below threshold hold. In my opinion
this way is better.
\vspace{0.3cm}

The linear response functional can be expressed as:
\be\lb{LinearResponseFunctDef2}
{\cal R}_{\op O}(\omega) = 
\sum_p\left(\bra{f_p(-\omega)}\op O\ket{\psi_p} + \bra{\psi_p}\op O\ket{f_p(\omega)}\right) \;.
\ee

@@

\vspace{0.3cm}
\noindent{TODO: Numerical example (convergence below threshold of the Benzene spectrum computed in different computational setup).}
